---
title: "Exploratory analysis"
author: "Leo"
date: "2025-11-28"
output: html_document
---

### Libraries

```{r libraries, include = FALSE}
library("corrplot")
library(dplyr)
library(MASS)
library(rgl)
library(DepthProc)
library(hexbin)
library(aplpack)
library(robustbase)
```

```{r dataset implementation, include=FALSE}
#dsMERGIATO <- read.csv("/Users/leomarcellopoli/Documents/Nonparametric/Project/Non-parametric-Statistics-2025-2026/dataset/dataset_MERGIATO", header = TRUE, sep = ",")
```

```{r set seed, include=FALSE}
set.seed(123) # Per riproducibilità
```

## Import dataset

```{r read dataset, include=FALSE}
indici_random <- sample(1:nrow(dsMERGIATO), 20000)
dsridotto <- dsMERGIATO[indici_random, -1]      # e senza date
dsclorofilla <- as.data.frame(dsridotto[, which(colnames(dsridotto) == "Chl")])
dscovariate <- dsridotto[, -which(colnames(dsridotto) == "Chl")]
```

## Correlation plot

```{r}
corrplot(cor(dsridotto), method = "circle")
```

## Scatterplot

```{r scatterplot, echo=TRUE}
for (i in 1:ncol(dscovariate)) {
  
  # A. Estrazione dinamica del nome della colonna X
  nome_covariata <- colnames(dscovariate)[i]
  
  # B. Plot con xlab dinamico
  # plot(X, Y, ...)
  plot(dscovariate[,i], 
       dsclorofilla[,1], 
       pch = 16, 
       cex = 0.5, # Punti piccoli
       col = rgb(0.7, 0, 0, 0.4), # Colore rosso trasparente
       ylab = "Clorofilla (Chl)",
       xlab = nome_covariata, # <--- QUI STA LA CORREZIONE
       main = paste("Chl vs", nome_covariata)
  ) 
}
```

### Hexagonal binning

to visualize data density

```{r hexagonal binning, echo=TRUE}
for (i in 1:ncol(dscovariate)) {
  
  # A. Estrazione dinamica del nome della colonna X
  nome_covariata <- colnames(dscovariate)[i]
  
  # B. Hexagonal binning
  bin = hexbin(dscovariate[,i], dsclorofilla[,1], xbins=30, xlab = nome_covariata, ylab="Chl")
  plot(bin, main="Hexagonal Binning")

}
```

Molti comportamenti particolari che vanno un attimo analizzati.

## Depth measures

```{r depth measures, echo=TRUE}
for (i in 1:ncol(dscovariate)) {
  #dP <- depthPersp(cbind(dscovariate[,i], dsclorofilla[,1]), depth_params = list(method='Tukey'), xlab = nome_covariata, ylab = "Chl")
  dP <- DepthProc::depthContour(cbind(dscovariate[,i], dsclorofilla[,1]), depth_params = list(method='Tukey'))
}
```

```{r other outliers, echo = TRUE}
for (i in 1:ncol(dscovariate)) {
  
  # A. Estrazione dinamica del nome della colonna X
  nome_covariata <- colnames(dscovariate)[i]
  
  # B. 
  aplpack::bagplot(cbind(dscovariate[,i], dsclorofilla[,1]), show.whiskers = F, main = "bagplot")
}
```

It may look like there are a lot of outliers, but it is a very high amount and also it seems that the pattern is quite uniform in the sense that the red points don't seem strange values to delete from the original data.

So no outliers detected.
