---
title: "Exploratory analysis"
author: "Leo"
date: "2025-11-28"
output: html_document
---

### **Libraries**

```{r libraries, include = T}
library("corrplot")
library(dplyr)
library(MASS)
library(rgl)
library(DepthProc)
library(hexbin)
library(aplpack)
library(robustbase)
library(ggplot2)
library(tidyverse)
library(roahd)
```

### **Import dataset**

```{r dataset implementation, include=T}
data = read.csv("../dataset/dataset_MERGIATO.", header = TRUE, sep = ",")
data = data[data$Lat >= 45, ]
```

```{r set seed, include=T}
set.seed(123) # Reproducibility

rand_indexes = sample(1:nrow(data), 20000)
data_reduced = data[rand_indexes, -1]      # Removed the first column (Date)
data_chl = as.data.frame(data_reduced[, which(colnames(data_reduced) == "Chl")])
data_cov = data_reduced[, -which(colnames(data_reduced) == "Chl")]
```


# **Multivariate data Exploration**

## **Correlation plots**

```{r corrplot}
corrplot(cor(data_reduced), method = "circle")
```

## **Scatterplots**

```{r scatterplot, echo=TRUE}
for (i in 1:ncol(data_cov)) {
  
  cov_name = colnames(data_cov)[i]
  
  plot(data_cov[,i], 
       data_chl[,1], 
       pch = 16, 
       cex = 0.5, # Small points
       col = rgb(0.7, 0, 0, 0.4), # Transparent red
       ylab = "Chl",
       xlab = cov_name,
       main = paste("Scatterplot Chl/", cov_name)
  ) 
}
```

## **Hexagonal binning**

```{r hexagonal binning, echo=TRUE}
for (i in 1:ncol(data_cov)) {
  
  cov_name = colnames(data_cov)[i]
  
  # Hexagonal binning
  bin = hexbin(data_cov[,i], data_chl[,1], xbins=30, xlab = cov_name, ylab="Chl")
  plot(bin,
       main = paste("Hexagonal Binning Chl/", cov_name))
}
```

Molti comportamenti particolari che vanno un attimo analizzati.

## **Depth measures analysis**

```{r depth measures, echo=TRUE}
for (i in 1:ncol(data_cov)) {
  cov_name = colnames(data_cov)[i]
  
  # Tukey depth contour
  dP <- DepthProc::depthContour(
    cbind(data_cov[,i], data_chl[,1]), 
    depth_params = list(method = 'Tukey')
  )

  ggplot2::labs( # Needed because depthContour is a ggplot-like object
    x = cov_name, 
    y = "Chl", 
    title = paste("Tukey Depth Contour Chl/", cov_name)
  )
}
```

```{r other outliers, echo = TRUE}
for (i in 1:ncol(data_cov)) {
  cov_name <- colnames(data_cov)[i]
  
  # Bagplot
  aplpack::bagplot(cbind(data_cov[,i], data_chl[,1]), show.whiskers = F, 
                   main = paste("Bagplot Chl/", cov_name), xlab = cov_name, ylab = "Chl")
}
```

It may look like there are a lot of outliers, but it is a very high amount and also it seems that the pattern is quite uniform in the sense that the red points don't seem strange values to delete from the original data.

So no outliers detected.

# **Functional Data Exploration (daily)**

```{r data preprocessing}
{#codice per trasformare data in numerico
data = data %>%
  mutate(
    Date = as.Date(Date),
    Date_numeric = as.numeric(Date - as.Date("2021-12-31")) # as.Date("2021-12-31") è lo 0
  ) %>%
  dplyr::select(Date_numeric, everything(), -Date)
}

time_grid = sort(unique(data$Date_numeric))
coordinates = unique(group_by(data[,2:3]))
vars_to_func <- c("Chl", "Temp", "Salinity", "MLD", "uo", "vo", "Current_Speed", "Solar_Flux", "Heat_Flux")
coord_reference <- data %>%
  distinct(Lon, Lat) %>%
  arrange(Lon, Lat) # Deve essere identico all'arrange usato nel pivot_wider!
world_map = map_data("world", region = c("Italy", "Croatia", "Slovenia"))
```

```{r functional data, include=T}
data_fun <- list()

for(var_name in vars_to_func) {
  message(paste("Working on: ", var_name, "..."))
  
  # Matrix of values
  data_matrix = data %>%
    
    dplyr::select(Lon, Lat, Date_numeric, all_of(var_name)) %>%
    pivot_wider(names_from = Date_numeric, values_from = all_of(var_name)) %>% # Time on columns
    arrange(Lon, Lat) %>%
    dplyr::select(-Lon, -Lat) %>% # Remove coordinate columns --> to obtain numerical matrix
    as.matrix()
  
  # fData object
  f_obj = fData(grid = time_grid, values = data_matrix)
  
  # Save fData object in the list
  data_fun[[var_name]] = f_obj
}
rm(var_name, data_matrix, f_obj)
gc()

message("Done!")
```
## **Plot of variables during the year**

```{r plot curve variabili, echo = TRUE}
for(var_name in vars_to_func) {
  plot(data_fun[[var_name]], main = paste("Tutte le curve funzionali di", var_name), ylab = var_name)
  # 2. Aggiunta linee verticali (Griglia temporale)
  abline(v = seq(151, 516, by = 7),  # Sequenza da 151 a 516 ogni 7
         col = "gray",               # Colore grigio
         lty = 2,                    # Tipo di linea (2 = tratteggiata/dashed)
         lwd = 0.5)                  # Spessore sottile
}
```

## **Functional Boxplot**

Finally, we can create functional boxplots for each variable to visualize the distribution of functional data.

```{r functional boxplot all variables, echo=TRUE}
for(var_name in vars_to_func) {
  invisible(fbplot(data_fun[[var_name]], main = paste("Boxplot di", var_name), ylab = var_name))
  # 2. Aggiunta linee verticali (Griglia temporale)
  abline(v = seq(151, 516, by = 7),  # Sequenza da 151 a 516 ogni 9
         col = "gray",               # Colore grigio
         lty = 2,                    # Tipo di linea (2 = tratteggiata/dashed)
         lwd = 0.5)                  # Spessore sottile
}
```

# **Functional Data Exploration (weekly)**
```{r preprocessing, echo=TRUE}

data_weekly = data %>%
  filter((Date_numeric - 151) %% 7 == 0) %>%
  
  mutate(Week_Number = ((Date_numeric - 151) / 7) + 1) %>%
  
  dplyr::select(Lon, Lat, Week_Number, everything(), -Date_numeric)

time_weekly_grid = sort(unique(data_weekly$Week_Number))


```

```{r functional data, include=T}
data_fun_weekly <- list()

for(var_name in vars_to_func) {
  message(paste("Working on: ", var_name, "..."))
  
  # Matrix of values
  data_matrix = data_weekly %>%
    
    dplyr::select(Lon, Lat, Week_Number, all_of(var_name)) %>%
    pivot_wider(names_from = Week_Number, values_from = all_of(var_name)) %>% # Time on columns
    arrange(Lon, Lat) %>%
    dplyr::select(-Lon, -Lat) %>% # Remove coordinate columns --> to obtain numerical matrix
    as.matrix()
  
  # fData object
  f_obj = fData(grid = time_weekly_grid, values = data_matrix)
  
  # Save fData object in the list
  data_fun_weekly[[var_name]] = f_obj
}
rm(var_name, data_matrix, f_obj)
gc()

message("Done!")
```

## **Plot of variables during the year**

```{r plot curve variabili, echo = TRUE}
for(var_name in vars_to_func) {
  plot(data_fun_weekly[[var_name]], main = paste("Tutte le curve funzionali di", var_name), ylab = var_name)
  # 2. Aggiunta linee verticali (Griglia temporale)
  abline(v = seq(151, 516, by = 7),  # Sequenza da 151 a 516 ogni 7
         col = "gray",               # Colore grigio
         lty = 2,                    # Tipo di linea (2 = tratteggiata/dashed)
         lwd = 0.5)                  # Spessore sottile
}
```

## **Functional Boxplot**

Finally, we can create functional boxplots for each variable to visualize the distribution of functional data.

```{r functional boxplot all variables, echo=TRUE}
for(var_name in vars_to_func) {
  invisible(fbplot(data_fun_weekly[[var_name]], main = paste("Boxplot di", var_name), ylab = var_name))
  # 2. Aggiunta linee verticali (Griglia temporale)
  abline(v = seq(151, 516, by = 7),  # Sequenza da 151 a 516 ogni 9
         col = "gray",               # Colore grigio
         lty = 2,                    # Tipo di linea (2 = tratteggiata/dashed)
         lwd = 0.5)                  # Spessore sottile
}
```